---
title: "Is Sweden smarter than the rest of Europe?"
output: 
  html_notebook: 
    theme: flatly
---

The corona virus crisis has overshadowed the life of many EU citizens for quite some time now, and will likely continue to do so for the coming months. The strict social distancing measures taken by many governments leave work places deserted, factories standing still and shops closed. However, one small country on the Northern edge of Europe is going it's own way to tackle the CoV pandemic and it is highly interesting how this experiment will turn out. I am part of that experiment.

----------

### Sweden goes its own way

Some European countries are recently showing signs that the spreading of the virus is slowing down. 

...

The Johns Hopkins institute is collecting world wide data on COVID-19 and shares it on github for public scrutiny and re-use. I cloned the repository so that I can access the most recent data and update if neccesary (the repo is updated every day 12 AM).

```{bash, eval = FALSE}
git clone https://github.com/CSSEGISandData/COVID-19.git
```


### Yet another visualization of CoV spread?

Unfortunately yes. There is hardly a phenomen that got more news coverage than COVID-19, so what's the point of it? One motivation is simply to see how the Swedish strategy will turn out in the end. In contrast to all the world maps and bubble charts, I will focus purely on Sweden, Germany, and some of their neighboring countries in the EU that have similar population size, population density, and health care systems (to some degree only, of course).


```{r, message = FALSE}
# attach packages
library(tidyverse)
library(lattice)
library(latticeExtra)
library(Rtools)
```

-----

The first step is to read the `*.csv` tables for confirmed cases and deaths from the Johns Hopkins University's data repository. It contains global time series of infections in a 'wide' format which is not nice to work with, so we need to reshape it to a 'long' format as it is more typical for data bases. The long format has all observations of one type in a single column, and all meta data (such as country, or date) in others.


```{r, message = FALSE}
df_world_cases <- read_csv("COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
df_world_death <- read_csv("COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")

# preview a table
head(df_world_cases[-1])

# remove slashes from column names
colnames(df_world_cases) <- colnames(df_world_cases) %>% gsub("/", "_", .)
colnames(df_world_death) <- colnames(df_world_death) %>% gsub("/", "_", .)

# remove greenland and faroe islands from data sets
df_world_cases <- df_world_cases %>% filter(!Province_State %in% c("Greenland", "Faroe Islands"))
df_world_death <- df_world_death %>% filter(!Province_State %in% c("Greenland", "Faroe Islands"))

# select only some countries of interest
countries = c("Sweden", "Finland", "Norway", "Denmark", "Germany", "Poland")
df_world_cases <- filter(df_world_cases, Country_Region %in% countries)[-1]
df_world_death <- filter(df_world_death, Country_Region %in% countries)[-1]

# reshape to long format using gather
# we gather only dates in a single new column
df_world_cases <- gather(df_world_cases, "date", "cases", "1_22_20":"4_7_20")
df_world_death <- gather(df_world_death, "date", "cases", "1_22_20":"4_7_20")

# now that confirmed cases and deaths are all in one column
# we might as well combine the two dfs in one
df_world <- bind_cols(
  df_world_cases, 
  rename(df_world_death["cases"], deaths = cases)
  )

# parse date-time entries
df_world <- mutate(
  df_world, 
  date = as.POSIXct(strptime(date, format="%m_%d_%y"))
  )

# preview after re-shaping to long format
head(df_world)
```

-----

Now that the data is in a handy format for multivariate plotting, we can start to compare cases and fatalities for the selected countries. If I would plot e.g. the confirmed cases on the Y axis on a linear scale, it would be hard to make out differences due to the exponential nature of infectious diseases (2 people infecting 2 more each = 4, then 8, 16, 32, 64, and so on). This is why it is better to plot on a logarithmic axis, here I chose log 10 (1 means 10^1 = 10, 2 = 100, 3 = 1000, 4 = 10,000 and so on). 

```{r, fig.width = 6.5, fig.height = 3.5}
plot_function = function(metric) {

  xyplot(log10(get(metric)) ~ date, 
    df_world %>% filter(date > "2020-02-15"), 
    groups = Country_Region, as.table = TRUE,
    par.settings = custom.lattice(),
    xlab = "", ylab = "log10 cases", main = metric,
    type = "l", lwd = 2, 
    scales = list(alternating = FALSE, x = list(rot = 25, cex = 0.7)),
    between = list(x = 0.5, y = 0.5),
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col.line = grey(0.9))
      panel.xyplot(x, y, ...)
      panel.key(..., points = FALSE)
    }
  )
}

print(plot_function("cases"), split = c(1,1,2,1), more = TRUE)
print(plot_function("deaths"), split = c(2,1,2,1))
grid::grid.text(label = date(), x = 0.15, y = 0.04, 
  gp = grid::gpar(col = grey(0.5), cex = 0.7))
```

It is quite clear that Germany is dominating the selection of countries in terms of cases and fatalities, while the other countries have lower numbers and also a lower slope of the curve, particularly for the cases. The flattening of the line for _all countries_ means that the disease is not spreading expomnentially anymore, but that the rate of new infections is slowing down. However, the main problem with this representation is that countries with a large population (such as Germany) will always appear on top the figure, although they might take efficient measures to  combat spreading of COVID-19.

A fairer representation is not only scaling the data logarithmically, but also normalizing number of cases by population size. This can be easily done by dividing `cases` and `deaths` by the new variable `population_size`. I simply added the number of inhabitants per country in millions, obtained from [wikipedia](https://en.wikipedia.org/wiki/List_of_European_countries_by_population).

```{r, fig.width = 6.5, fig.height = 3.5, warning = FALSE}
# add new variable pop size
df_world <- df_world %>% 
  mutate(population_size = recode(Country_Region,
    Denmark = 5.787725,
    Finland = 5.539832,
    Germany = 83.832481,
    Norway = 5.411798,
    Poland = 37.860731,
    Sweden = 10.088474
    )
  )

# normalize cases and deaths per million of inhabitants
df_world <- mutate(df_world,
    cases_per_m = cases/population_size,
    deaths_per_m = deaths/population_size
  )

print(plot_function("cases_per_m"), split = c(1,1,2,1), more = TRUE)
print(plot_function("deaths_per_m"), split = c(2,1,2,1))
grid::grid.text(label = date(), x = 0.15, y = 0.04, 
  gp = grid::gpar(col = grey(0.5), cex = 0.7))
```


